On existence of integral point sets and their diameter bounds \footnote{ This work was carried out at Voronezh State University and supported by the Russian Science Foundation grant 19-11-00197. }

### 2024-06-27 char-4k-bound/char-4k-bound.tex
[4k3] WIP - sets of integers

@@

\section{Collinear subsets of integral point sets}
Integral point sets with many collinear points are very dominating examples.
In this section we estimate the cardinality of an intersection of an integral point set with a line segment.

\subsection{General results}

Assuming that the planar integral point set contains many collinear points,
the following result holds.
\begin{theorem}~\cite[Theorem 4]{kurz2008minimum}
	For $\delta > 0$, $\varepsilon > 0$, and $P\in\mathfrak{M}(2,n)$ with
	at least $n^\delta$ collinear points there exists a $n_0 (\varepsilon)$
	such that for all $n \geq n_0 (\varepsilon)$ we have
	\begin{equation}
		\operatorname{diam} P \geq n^{\frac{\delta}{4 \log 2(1+\varepsilon)}\log \log n}
		.
	\end{equation}
\end{theorem}
However, the estimate is rather unsuitable for our needs,
as it does not provide the values of all the constants.
To obtain the required estimate, we now prove a generalization of~\cite[Lemma 3]{our-vmmsh-2018-translit}.

\begin{definition}
	For a line segment $M_1 M_2$ and an integer $k$, such that $-|M_1 M_2| < k < |M_1 M_2|$,
	we define a $\rho(k,M_1, M_2)$-curve as the set of points $N$
	for which the equality $|N M_1| - |N M_2| = k$ holds.
\end{definition}
So, in the planar case a $\rho(k,M_1, M_2)$-curve
is a branch of a hyperbola for $k\neq 0$ and the perpendicular bisector of line segment $M_1 M_2$ for $k=0$.

\begin{proposition}
	\label{obs:rho_curves}
	If points $M_1,M_2,M_3,M_4$ are situated on a straight line,
	the equality $|M_1 M_2| = |M_3 M_4| =h $ holds, line segments $M_1 M_2$ and $M_3 M_4$ do not coincide
	and vectors $\overrightarrow{M_1 M_2}$ and $\overrightarrow{M_3 M_4}$ are of equal direction,
	then for a fixed integer $k$, $|k| < h$, the $\rho(k,M_1, M_2)$-curve and the $\rho(k,M_3, M_4)$-curve
	do not intersect.
\end{proposition}

\begin{proof}
	For $k=0$, the claim is obvious.
	Indeed, in this case the $\rho(k,M_1, M_2)$-curve and the $\rho(k,M_3, M_4)$-curve
	are distinct hyperplanes orthogonal to $l$ and thus do not intersect.

	For $k\neq 0$, suppose the contrary.
	Let $N$ be a point in the intersection.

	Consider a plane $\Pi \supset (l \cup \{N\})$
	with a Cartesian coordinate system such that $M_i = (x_i,0)$ for $1 \leq i \leq 4$.
	Let $R_1$ and $R_2$ be the projections
	of the $\rho(k,M_1, M_2)$-curve and the $\rho(k,M_3, M_4)$-curve onto $\Pi$.

	Then $R_1$ is a branch of a hyperbola defined by the equation
	\begin{equation}
		x- \frac{x_1+x_2}{2}=a(h,k)\frac{\sqrt{y^2+b^2(h,k)}}{b(h,k)}
	\end{equation}
	and $R_2$ is a branch of a hyperbola defined by the equation
	\begin{equation}
		x- \frac{x_3+x_4}{2}=a(h,k)\frac{\sqrt{y^2+b^2(h,k)}}{b(h,k)}
		,
	\end{equation}
	where $a(h,k)$ and $b(h,k)$ are coefficients which depend on $h$ and $k$.
	For $N=(x_N, y_N) \in R_1 \cap R_2$ we have
	\begin{equation}
		\begin{cases}
			x_N- \dfrac{x_1+x_2}{2}=a(h,k)\dfrac{\sqrt{y_N^2+b^2(h,k)}}{b(h,k)}
			\\[4pt]\\
			x_N- \dfrac{x_3+x_4}{2}=a(h,k)\dfrac{\sqrt{y_N^2+b^2(h,k)}}{b(h,k)}
		\end{cases}
	\end{equation}
	and, thus, $x_1+x_2 = x_3+x_4$.
	On the other hand, $x_2 - x_1 = x_4 - x_3 = h$,
	and we conclude that $x_1 = x_3$ as well as $x_2 = x_4$.
	So, line segments $M_1 M_2$ and $M_3 M_4$ coincide.
	The contradiction concludes the proof.
\end{proof}

\begin{lemma}
	\label{lem:2k-1_segments}
	Let $M \in \mathfrak{M}(2,\mathbb{N})$ and let $l$ be a straight line.
	Then for every $k\in\mathbb{N}$ there are at most $2k-1$ segments $\Delta_i \subset l$ with endpoints in $M$,
	such that $|\Delta_i| = k$.
\end{lemma}
\begin{proof}
	Consider a point $N\in M \setminus l$.
	Let $\Delta_i = M_i^1 M_i^2$, and
	let vectors $\overrightarrow{M_i^1 M_i^2}$ and $\overrightarrow{M_j^1 M_j^2}$ be of equal direction for every $i$ and $j$.
	Then for each $i$ there is a $\rho(n_i, M_i^1, M_i^2)$-curve containing $N$.
	Due to Proposition~\ref{obs:rho_curves}, all $n_i$ are distinct;
	otherwise the $\rho(n_i,M_i^1, M_i^2)$-curve and the $\rho(n_i,M_j^1, M_j^2)$-curve, $j\neq i$, do not intersect.
	There can be only $2k-1$ distinct values for $n_i$,
	so there are at most $2k-1$ distinct segments $\Delta_i$.
\end{proof}

Setting $k=1$ in Lemma~\ref{lem:2k-1_segments},
we obtain the following result (which is exactly~\cite[Lemma 3]{our-vmmsh-2018-translit}).
\begin{corollary}
	\label{cor:only_one_distance_1_on_straight_line}
	Let $M \in \mathfrak{M}(2,\mathbb{N})$ and let $l$ be a straight line.
	Then there is at most one pair of points $M_1,M_2\in M \cap l$
	such that $|M_1 M_2| = 1$.
\end{corollary}

\begin{lemma}
	\label{lem:line_segment_with n_squared_plus_one_points}
	Let $\Delta$ be a straight line segment, $|\Delta|=l$ and $M \in \mathfrak{M}(2,\mathbb{N})$.
	Let $\#(\Delta \cap M) = n^2 + 1$.
	Then
	\begin{equation}
		l \geq \frac{2}{3}n^3+\frac{1}{2}n^2-\frac{1}{6}n
		.
	\end{equation}
\end{lemma}

\begin{proof}
	Any $n^2+1$ points, including the endpoints of $\Delta$, split the segment $\Delta$ into $n^2$
	sequential segments $\Delta_i$.
	Due to Lemma~\ref{lem:2k-1_segments}, there is at most one segment of length 1,
	at most three segments of length 2, etc.
	The following two expressions for sums conclude the proof:
	\begin{equation}
		1 + \sum_{k=1}^n (2k-1) = n^2 + 1
		,
	\end{equation}
	\begin{equation}
		\sum_{k=1}^n k(2k-1) = \frac{2}{3}n^3+\frac{1}{2}n^2-\frac{1}{6}n
		.
	\end{equation}
\end{proof}

Now we will estimate the length of a line segment that intersects an integral point set
by an arbitrary number of points.

\begin{lemma}
	\label{lem:line_segment_cardinality}
	Let $\Delta$ be a straight line segment, $|\Delta| = b$ and $M \in \mathfrak{M}(2,\mathbb{N})$.
	Let $\#(\Delta \cap M) = t$.
	Then
	\begin{equation}
		\label{eq:estimate_for_segment_length}
		b\geq \frac{2}{3}t^{3/2}-\frac{3}{2}t+\frac{5}{6}t^{1/2}
		.
	\end{equation}
\end{lemma}

\begin{proof}
	Let $f(k)$ denote the minimal length of a line segment
	that intersects an integral point set by $k$ points.
	We observe that $f(k) > f (k-1)$.
	Due to Lemma~\ref{lem:line_segment_with n_squared_plus_one_points},
	$f(n^2+1) \geq \frac{2}{3}n^3+\frac{1}{2}n^2-\frac{1}{6}n$.

	%Consider $t$ such that $n^2 + 1 \leq t \leq (n+1)^2$. %,
	%then $f(t) > f(n^2 +1)$.% and $n \geq \sqrt{t} - 1$.
	For $t\in\mathbb{N}$ the inequality $(\sqrt{t} - 1)^2 +1 \leq t$ holds,
	thus
	%and derive that
	\begin{equation}
		\label{eq:line_segment_length}
		f(t) \geq f((\sqrt{t} - 1)^2 +1) \geq \frac{2}{3}(\sqrt{t} - 1)^3+\frac{1}{2}(\sqrt{t} - 1)^2-\frac{1}{6}(\sqrt{t} - 1)
		=
		\frac{2}{3}t^{3/2}-\frac{3}{2}t+\frac{5 }{6}t^{1/2}
		.
	\end{equation}
\end{proof}

\subsection{Special results}

In~\cite{kurz2008bounds}, thorough investigation of planar integral point sets with diameter at most 10~000
is carried out;
thus, we can restrict our attention to intersections of planar integral point sets with line segments of length more than 10~000.

Lemma~\ref{lem:line_segment_cardinality} leads to the following proposition.

\begin{proposition}
	\label{obs:estimate_points_on_straight_line}
	Let $\Delta$ be a straight line segment, $|\Delta| = b$ and $M \in \mathfrak{M}(2,\mathbb{N})$.
	Let $\#(\Delta \cap M) = k$ and $b>10000$.
	Then $k \leq \gamma_2 b + 6$,
	where
	\begin{equation}
		\gamma_2 = \frac{3846}{2593 \sqrt{647}-5823} = 0.063958...
	\end{equation}
\end{proposition}

\begin{proof}
	We know the maximum number of points for planar integral point sets of diameters at most $10000$.
	Thus, we are interested in integral $t$ such that estimate~\eqref{eq:estimate_for_segment_length}
	holds for $b > 10000$.
	So, let us consider $t\geq 647$ and find a coefficient $\gamma_2$,
	such that the inequality
	\begin{equation}
		\label{eq:inequality_for_linear_estimate_of_segment_length}
		\frac{2}{3}t^{3/2}-\frac{3}{2}t+\frac{5 }{6}t^{1/2} \geq \frac{t-6}{\gamma_2}
	\end{equation}
	holds for all $t\geq 647$.
	For such $t$,
	the left-hand side of~\eqref{eq:inequality_for_linear_estimate_of_segment_length} obviously grows faster than the right-hand side.
	Turning~\eqref{eq:inequality_for_linear_estimate_of_segment_length} into the same equality
	and solving it for $t=647$, we obtain the required estimate.
\end{proof}

To conclude, we remark that
Corollary~\ref{cor:only_one_distance_1_on_straight_line}
and
Proposition~\ref{obs:estimate_points_on_straight_line}
are the only results of this section
that are used in the proofs below;
all the other are auxiliary.

\section{Planar integral point set in a square}

In this section, we develop some methods that employ the fact that every planar integral point set is situated in a square.
Then we establish a connection between distances in integral point sets and the Point Packing in the Square problem.

\subsection{General results}

Let us introduce the problem, basic notions and results.

\begin{problem}[Point Packing in a Square (PPS)~\cite{locatelli2002packing,costa2013valid}]
	\label{problem:PPS}
	Given an integer $k > 1$, place $k$ points in the unit square $U = [ 0 , 1 ]^2$ such that their
	minimum pairwise distance $m$ is maximal.
\end{problem}

\begin{definition}
	For each $k > 1$, the corresponding maximal distance $m$ from Problem~\ref{problem:PPS}
	is called the $k$-th PPS coefficient and denoted by $\varphi_k$.
\end{definition}
So, it's impossible to place $k$ points in a unit square in such a way that each pairwise distance of the points is greater than $\varphi_k$.

\begin{theorem}
	\label{thm:varphi_k_bounds}
	\cite{costa2013valid}
	For every $k\geq 2$ the following inequality holds:
	\begin{equation*}
		\sqrt{\frac{2}{k\sqrt{3}}}
		\leq
		\varphi_k
		\leq
		\frac{1}{k-1} +
		\sqrt{
			\frac{1}{(k-1)^2}
			+
			\frac{2}{(k-1)\sqrt{3}}
		}
	\end{equation*}
\end{theorem}

Now we will estimate the side length of a square which contains a given planar integral points set.

\begin{lemma}
	\cite[Lemma 4]{our-vmmsh-2018-translit}
	\label{lem:square_container}
	Let $M\in\mathfrak{M}(2,n)$, $\operatorname{diam} M = d$.
	Then $M$ is situated in a square of side length $d$.
\end{lemma}

\begin{proof}
	Consider $M_1, M_2 \in M$ such that $|M_1 - M_2| = d$
	and a Cartesian coordinate system such that
	$M_1 = (-d/2; 0)$, $M_2 = (d/2; 0)$.

	Then for $M_i =(x_i, y_i)\in M$, $i > 2$,
	the inequality $|x_i| < d/2$ holds
	(otherwise we get $|M_i - M_1| > d$ or $|M_i - M_2| > d$).
	Moreover, $|y_i| < d$ (otherwise we get $|M_i - M_1| > d$ or $|M_i - M_2| > d$).

	Set $y_{\max} = \max_{i} y_i$, $y_{\min} = \min_{i} y_i$, $M_{\max}=(x_{\max}, y_{\max})$, $M_{\min}=(x_{\min}, y_{\min})$
	(in case of multiple points providing the maximum or the minimum, the choice is arbitrary).
	Then
	\begin{multline}
		d \geq |M_{\max} - M_{\min}| = \sqrt{(x_{\max} - x_{\min})^2 + (y_{\max} - y_{\min})^2}
		\\\geq
		\sqrt{(y_{\max} - y_{\min})^2} =
		y_{\max} - y_{\min}
	\end{multline}
	Thus, the sides of the square that contains $M$ are $x=\pm d/2$,
	$y=y_{\max}$ and $y=y_{\max} - d$.
\end{proof}

\subsection{Special results}

It is clear that for $n>123$ we have $d(2,n) \geq d(2,123)$.
Based on the exact values of $d(2,n)$ for $ 3 \leq n\leq 122$ and the estimate $d(2,123)>10000$
\cite{kurz2008bounds}, we derive the following proposition.
\begin{proposition}
	\label{obs:4_leq_n_leq_21491}
	The inequality
	\begin{equation}
		d(2,n) \geq 3^{1/4}\cdot2^{-3/2} \cdot n
	\end{equation}
	holds for $4 \leq n \leq 21491$.
\end{proposition}

Therefore, we will focus on planar integral points sets of more than 21491 points.

Now we will use the upper bound of Theorem~\ref{thm:varphi_k_bounds}
to estimate $\varphi_n$ with $\beta/\sqrt{n-2}$.

\begin{proposition}
	\label{varphi_n_where_n_geq_21492}
	For $n \geq 21492$ we have
	\begin{equation}
		\varphi_n \leq \varphi_{n-1} \leq \frac{\beta}{\sqrt{n-2}}
		,
	\end{equation}
	where
	\begin{equation}
		\beta = \frac{1}{\sqrt{21490}} + \sqrt{ \frac{2}{\sqrt{3}} + \frac{1}{21490} } < 1.07464
		.
	\end{equation}
\end{proposition}

\begin{proof}
	Indeed, for $n \geq 21492$ we have
	\begin{align*}
		\varphi_n
		\leq
		\varphi_{n-1}
		\leq
		\frac{1}{n-2} +
		&
		\sqrt{
			\frac{1}{(n-2)^2}
			+
			\frac{2}{(n-2)\sqrt{3}}
		}
		\\&=
		\frac{1}{\sqrt{n-2}}
		\cdot
		\frac{1}{\sqrt{n-2}}
		 +
		\sqrt{
			\frac{1}{(n-2)^2}
			+
			\frac{2}{(n-2)\sqrt{3}}
		}
		\\&\leq
		\frac{1}{\sqrt{n-2}}
		\cdot
		\frac{1}{\sqrt{21490}}
		 +
		\sqrt{
			\frac{1}{(n-2)^2}
			+
			\frac{2}{(n-2)\sqrt{3}}
		}
		\\&=
		\frac{1}{\sqrt{n-2}}
		\left(
		\frac{1}{\sqrt{21490}}
		 +
		\sqrt{
			\frac{1}{n-2}
			+
			\frac{2}{\sqrt{3}}
		}
		\right)
		\\&\leq
		\frac{1}{\sqrt{n-2}}
		\left(
		\frac{1}{\sqrt{21490}}
		 +
		\sqrt{
			\frac{1}{21490}
			+
			\frac{2}{\sqrt{3}}
		}
		\;
		\right)
		.
	\end{align*}
\end{proof}

\section{Lower bound for the diameter}

In this section,
we improve the lower bound for the minimum diameter of planar integral point sets
employing the results of two previous sections.

\subsection{General results}

\begin{definition}
	A \textit{cross} for points $M_1$ and $M_2$, denoted by $cr(M_1,M_2)$, is the union of two straight lines:
	the line through $M_1$ and $M_2$,
	and the perpendicular bisector of line segment $M_1 M_2$.
\end{definition}

\begin{lemma}
	\label{lem:intervals_cross}
	If open line segments $M_1 M_2$ and $M_3 M_4$ do not intersect,
	%$M_1 \neq M_2$, $M_3 \neq M_4$,
	then the set $cr(M_1,M_2) \cap cr(M_3,M_4)$ is either a straight line or contains 2 or 4 points.
\end{lemma}

\begin{proof}
	Consider straight lines $l^1_1$, $l^1_2$, $l^2_1$, $l^2_2$
	such that $cr(M_1,M_2) = l^1_1 \cup l^1_2$ and $cr(M_3,M_4) = l^2_1 \cup l^2_2$.

	The cases $l^1_1 = l^2_1$, $l^1_2 = l^2_2$ and $l^1_1 = l^2_2$, $l^1_2 = l^2_1$ are impossible,
	because the centers of open line segments $M_1 M_2$ and $M_3 M_4$ do not coincide as far as
	the open line segments do not intersect.
	So, all the possible cases (up to enumeration of the straight lines) are:
	\begin{itemize}
		\item
			$l^1_1 = l^2_1$, $l^1_2 \neq l^2_2$.
			The straight lines $l^1_2$ and $l^2_2$ are both orthogonal to $l^1_1$ and thus do not intersect;
			thus, $cr(M_1,M_2) \cap cr(M_3,M_4) = l^1_1 = l^2_1$.
		\item
			$l^1_1$ is parallel to $l^2_1$ and $l^1_2$ is parallel to $l^2_2$.
			Then
			\begin{equation*}
				cr(M_1,M_2) \cap cr(M_3,M_4) = (l^1_1 \cup l^1_2)\cap(l^2_1 \cup l^2_2) =
				(l^1_1 \cap l^2_2) \cup (l^1_2 \cap l^2_1)
				,
			\end{equation*}
			that is two points.
		\item
			$l^1_1$ is neither parallel nor orthogonal to $l^2_1$.
			Then the intersection contains 4 points.
	\end{itemize}
\end{proof}

\begin{remark}
	It is important that we consider the intersection of open line segments in the claim of Lemma~\ref{lem:intervals_cross},
	so e.g. the cases $M_1 = M_3$ and $M_3 \in M_1 M_2$ satisfy the conditions of Lemma~\ref{lem:intervals_cross}.
\end{remark}

\begin{lemma}
	\label{lemma:quadr_diag_edges}
	Let $ABCD$ be a convex quadrilateral on the plane.
	Then $\max\{AC,BD\}>\min\{AB,BC,CD,DA\}$,
	that is at least one diagonal is greater than at least one side.
\end{lemma}

\begin{proof}
	Suppose the contrary.
	Then $AC$ in the minimal side of the triangle $ABC$ and, thus, the angle $ABC$ is acute.
	So are also angles $BCD$, $CDA$ and $DAB$.
	We conclude that all angles of a convex quadrilateral $ABCD$ are acute, which is impossible.
	The contradiction concludes the proof.
\end{proof}

\subsection{Special results}

Now we are ready to prove the main theorem of the section.

\begin{theorem}
	\label{thm:main_estimate}
	If $n\geq 4$, then $d(2,n) \geq \gamma (n - 2)$,
	where
	\begin{multline}
		0.46530... =
		3^{1/4} \cdot 2^{-3/2}
		\\ >
		\gamma = \frac{\sqrt{16 {{\left( \sqrt{\frac{2}{\sqrt{3}}+\frac{1}{21490}}+\frac{1}{\sqrt{21490}}\right) }^{2}}+\frac{14791716}{{{\left( 2593 \sqrt{647}-5823\right) }^{2}}}}-\frac{3846}{2593 \sqrt{647}-5823}}{8 {{\left( \sqrt{\frac{2}{\sqrt{3}}+\frac{1}{21490}}+\frac{1}{\sqrt{21490}}\right) }^{2}}}
		\\
		> 0.45557
		> \frac{5}{11}
		.
	\end{multline}
\end{theorem}

\begin{proof}
	For $4 \leq n \leq 21491$, the assertion of the theorem follows immediately from Proposition~\ref{obs:4_leq_n_leq_21491}.
	Let us consider $M\in \mathfrak{M}(2,n)$, $n \geq 21492$, $\operatorname{diam} M = b$.
	Lemma~\ref{lem:square_container} yields that $M$ is situated in a square of side length $b$.
	Let $M_1, M_2, M_3, M_4$ be points of $M$ such that the distances $|M_1 M_2|$ and $|M_3 M_4|$
	are minimal in $M$ ($M_2$ and $M_3$ may coincide).
	Then $|M_1 M_2| \leq b\varphi_{n-1}$ and $|M_3 M_4| \leq b\varphi_{n-1}$.
	Due to Lemma~\ref{lemma:quadr_diag_edges}, open line segments $M_1 M_2$ and $M_3 M_4$ do not intersect
	(otherwise they are not minimal).

	Let $C = cr(M_1 M_2) \cap cr(M_3 M_4)$.
	Each point $N\in M$ satisfies one of the following conditions:

	a) $N$ belongs to $C$; overall at most $\gamma_2 b + 6$ points by Lemma~\ref{lem:intervals_cross} and Proposition~\ref{obs:estimate_points_on_straight_line};

	b) $N$ belongs to the intersection of one of $|M_1 M_2| - 1$ hyperbolas
	with one of $|M_3 M_4| - 1$ hyperbolas; overall at most $4 (|M_1 M_2| - 1)(|M_3 M_4| - 1)$ points;

	c) $N$ belongs to the intersection of one of $|M_1 M_2| - 1$ hyperbolas with $cr(M_3 M_4)$;
	overall at most $4 (|M_1 M_2| - 1)$ points;

	d) $N$ belongs to the intersection of one of $|M_3 M_4| - 1$ hyperbolas with $cr(M_1 M_2)$;
	overall at most $4 (|M_3 M_4| - 1)$ points;
	\\
	(see~\cite{erdos1945integral} for details).
	Summing up the above cases, we obtain the following estimate:
	\begin{equation}
		\label{eq:estimate_n_varphi_n_squared}
		n \leq 4 b^2 \varphi_{n-1}^2 - 4 + \gamma_2 b + 6
		.
	\end{equation}
	Proposition~\ref{varphi_n_where_n_geq_21492} turns estimate~\eqref{eq:estimate_n_varphi_n_squared}
	into the following:
	\begin{equation}
		\label{eq:estimate_n_sqrt_n-2}
		n-2 \leq 4 b^2 \frac{\beta^2}{n-2} + \gamma_2 b
		,
	\end{equation}
	which obviously leads to inequality
	\begin{equation}
		1 \leq 4\beta^2 \left(\frac{b}{n-2}\right)^2  + \gamma_2 \frac{b}{n-2}
		.
	\end{equation}
	Let us denote $\lambda = b/(n-2)$ and solve the following quadratic inequality for $\lambda$:
	\begin{equation}
		\label{eq:square_inequality_lambda}
		4\beta^2 \lambda^2  + \gamma_2 \lambda - 1 \geq 0
		.
	\end{equation}
	The discriminant is $\gamma_2^2 + 16 \cdot \beta^2$,
	so we obtain the following estimate:
	\begin{equation}
		\frac{b}{n-2} = \lambda \geq \frac{-\gamma_2 + \sqrt{\gamma_2^2 + 16 \cdot \beta^2}}{8\beta^2}
		.
	\end{equation}
	Calculating the expression above for our $\beta$ and $\gamma_2$, we conclude the proof.
\end{proof}

\begin{corollary}
	If $n\geq 4$, then $d(2,n) > \frac{5}{11} n$.
\end{corollary}

\begin{proof}
	For $4 \leq n \leq 21491$, the claim follows from Proposition~\ref{obs:4_leq_n_leq_21491} immediately.
	For $n > 21491$, the inequality $0.45557(n-2) > \frac{5}{11}n$ holds.
\end{proof}

We can improve the constant in Theorem~\ref{thm:main_estimate},
if some new exact values of $d(2,n)$ are found;
however, the obtained constant is bounded by $3^{1/4}\cdot2^{-3/2}$ from above.
More precisely, we have the following theorem.

\begin{theorem}
	For every $\varepsilon > 0$ there exists a number $n_0$ such that the inequality
	\begin{equation}
		d(2,n) \geq n\cdot(3^{1/4}\cdot 2^{-3/2} - \varepsilon)
	\end{equation}
	holds for every $n>n_0$,
	that is
	\begin{equation}
		\lim_{n\to\infty} \frac{d(2,n)}{n} \geq 3^{1/4}\cdot 2^{-3/2}
		.
	\end{equation}
\end{theorem}

\section{Integral point sets with distance 1}

\subsection{Constructing planar integral point sets with distance 1}

\begin{definition}
	A set $M\in\mathfrak{M}(2,n)$ is called \textit{facher}
	if $M$ consists of $n-1$ points on a straight line
	and one point out of the line.
\end{definition}

\begin{definition}
	A set $M\in\mathfrak{M}(m,n)$ is called \textit{optimal}
	if $\operatorname{diam}M=d(m,n)$.
\end{definition}

\begin{definition}
	\cite{kurz2005characteristic}
	A squarefree number $q$ is called the \textit{characteristic} of $M\in\mathfrak{M}(2,\mathbb{N})$,
	if for any points $M_1, M_2, M_3 \in M$ the area of triangle $M_1 M_2 M_3$
	is $p_{1,2,3}\sqrt{q}$ for some rational $p_{1,2,3}$.
\end{definition}
For a given $M\in\mathfrak{M}(2,\mathbb{N})$, the characteristic is determined uniquely.

Facher sets are the simplest planar integral point sets.
It is known that for $9\leq n \leq 122$ all the optimal sets are facher~\cite{kurz2008minimum}.
For every cardinality $n$ and every squarefree number $q$
there exists a facher set $M\in\mathfrak{M}(2,n)$ with characteristic $q$~\cite[Theorem 5]{our-vmmsh-2018-translit}.
In~\cite{antonov2008maximal}, the facher sets of characteristic 1 were investigated; they were called \textit{semi-crabs}.

For every integer $n\geq 3$ Solymosi presented~\cite{solymosi2003note} a construction of a facher integral point set
$M\in\mathfrak{M}(2,n)$
such that equality $|M_1 M_2| = 2$ holds for some $M_1, M_2 \in M$.
The constructed set has both odd and even distances.

Now we improve Solymosi's result.

\begin{construction}
	\label{con:planar_set_with_minimeter_1}
	Let us choose a positive integer $k > 1$ and set
	\begin{equation}
		a = 2^{2^k} - 1
		.
	\end{equation}
	Then
	\begin{equation}
		\label{eq:a_equiv_3_mod_4}
		a \equiv 3 \mod 4
	\end{equation}
	and, moreover,
	\begin{multline}
		a = \left(2^{2^{k-1}}\right)^2 - 1
		=
		\left(2^{2^{k-1}} + 1\right) \left(2^{2^{k-1}} - 1\right)
		\\=
		\left(2^{2^{k-1}} + 1\right) \left(2^{2^{k-2}} + 1\right) \cdot ... \cdot \left(2^{2^1} + 1\right) \left(2^2 - 1\right)
		.
	\end{multline}
	We set $d_j = 2^{2^j} + 1$ for $1 \leq j \leq k-1$.
	Then $d_j \equiv 1 \mod 4$.

	Let $c_J = \prod_{j\in J} d_j$ for every subset of indices $J\subset I = \{1,2,...,k-1\}$
	(and  $c_\varnothing = 1$).
	We obtain
	\begin{equation}
		\label{eq:c_J_equiv_1_mod_4}
		c_J\equiv 1 \mod{4}
	\end{equation}
	and, moreover, $a$ is divisible by $c_J$.

	Statements~\eqref{eq:a_equiv_3_mod_4} and~\eqref{eq:c_J_equiv_1_mod_4} yield $a/c_J \equiv 3 \mod 4$.
	Let $b_J = (c_J - a/c_J)/2$, then $b_J \equiv 1 \mod 2$.
	Let us further take $g_J = (c_J + a/c_J)/2$, then $g_J \equiv 0 \mod 2$.

	Next we define the coordinates of the points as following:
	\begin{equation}
		M_{J\pm} =\left(\pm\frac{b_J}{2}, 0\right)
		,
		~~
		N   =\left(0, \frac{\sqrt{a}}{2}\right)
		.
	\end{equation}
	Then the distances are:
	\begin{multline}
		|N M_{J\pm}|
		=
		\left(\frac{b_J^2}{4} + \frac{a}{4}\right)^{1/2}
		=
		\frac{1}{2}\left(\left(\frac{c_J - a/c_J}{2}\right)^2 + a\right)^{1/2}
		\\=
		\frac{1}{2}\left( \left(\frac{c_J}{2}\right)^2 - \frac{a}{2} + \left(\frac{a}{2c_J}\right)^2 + a\right)^{1/2}
		=
		\frac{1}{2}\left( \left(\frac{c_J}{2}\right)^2 + \frac{a}{2} + \left(\frac{a}{2c_J}\right)^2    \right)^{1/2}
		\\=
		\frac{1}{2}\left(\left(\frac{c_J + a/c_J}{2}\right)^2\right)^{1/2}
		=
		\frac{1}{2}\left(\frac{c_J + a/c_J}{2}\right)
		=
		\frac{g_J}{2}
		\in\mathbb{N}
		,
	\end{multline}
	\begin{equation}
		|M_{J_1 \pm}  M_{J_2 \pm}|
		=
		\left|\frac{b_{J_1}}{2} \pm \frac{b_{J_2}}{2}\right|
		=
		\left|\frac{b_{J_1} \pm b_{J_2}}{2}\right|
		\in\mathbb{N}
		.
	\end{equation}
	In particular, for $H = \{k-1\}$ we obtain $C_H = 2^{2^{k-1}}+1$ and
	\begin{equation}
		b_H =
		\left( 2^{2^{k-1}}+1 - \frac{a}{2^{2^{k-1}}+1} \right)/2
		=
		\left(2^{2^{k-1}}+1 - \left(  2^{2^{k-1}}-1 \right) \right)/2
		=
		1
		.
	\end{equation}
	Thus, one of the distances is
	\begin{equation}
		|M_{H+}  M_{H-}|
		=
		\left|\frac{b_{H}}{2} - \frac{-b_{H}}{2}\right|
		=
		\left|\frac{1}{2} - \frac{-1}{2}\right|
		= 1
		.
	\end{equation}

	Note that all the points $M_{J\pm}$ are distinct:
	the equality $b_J =  b_K$ implies $J=K$;
	the equality $b_J = -b_K$ implies $c_J = -c_K$ or $c_J = a / c_K$.
	The first case is impossible because both $c_J$ and $c_K$ are positive;
	the second case contradicts~\eqref{eq:a_equiv_3_mod_4} and~\eqref{eq:c_J_equiv_1_mod_4}.

	So, $M = \{ M_{J\pm}, N\}$ is indeed a planar integral point set of $2^k+1$ points,
	and distance 1 occurs in $M$.
	Since $k$ can be taken arbitrary large, it follows that we can construct
	a planar integral point set of arbitrary large cardinality so that the distance 1 occurs in that set.
\end{construction}

\begin{remark}
	\label{rem:planar_set_with_minimeter_1_of_3_points}
	Applying Construction~\ref{con:planar_set_with_minimeter_1} to $k=1$ ad litteram,
	we get $a = 3$, $I=\varnothing$, $b_\varnothing = -1$ and then
	obtain an equilateral triangle of side length 1.
	This triangle is obviously the optimal set in $\mathfrak{M}(2,3)$.
\end{remark}

\begin{remark}
	\label{rem:planar_set_with_minimeter_1_of_4_and_5_points}
	For $k=2$ in Construction~\ref{con:planar_set_with_minimeter_1},
	we obtain the optimal set in $\mathfrak{M}(2,5)$ presented in~\cite[Fig. 1]{harborth1993upper}.
	If we remove one point from it,
	then we get one of the two optimal sets in $\mathfrak{M}(2,4)$.
\end{remark}

\subsection{Description of planar integral point sets with distance 1}

%Now we need two more definitions.
\begin{definition}
	\cite{antonov2008maximal}
	A set $M\in \mathfrak{M}(m,n)$ is \textit{maximal},
	if there is no set $M'\in \mathfrak{M}(m,n+1)$
	such that $M \subsetneq M'$.
\end{definition}

%\begin{definition}
%	A set $M\in \mathfrak{M}(m,n)$ is \textit{submaximal},
%	if there is only one maximal integral point set $M'\in \mathfrak{M}(m,n+1)$,
%	such that $M \subset M'$.
%\end{definition}

In order to describe all the sets $M \in \mathfrak{M}(2,n)$ with distance 1,
we need~\cite[Proposition 6]{our-vmmsh-2018-translit}.
For the reader's convenience, we will state a slightly rephrased version of it and provide the proof.
\begin{lemma}
	\label{lem:no_4_points_in_semigeneral_position_with_distance_1}
	Let $M=\{M_1,M_2,M_3,M_4\}\in \mathfrak{M}(2,4)$ such that $|M_1 M_2|=1$.
	If $l$ is the straight line through points $M_1$ and $M_2$,
	then one of the points $M_3$ or $M_4$ belongs to $l$.
\end{lemma}

\begin{proof}
	Let $m$ denote the perpendicular bisector of line segment $M_1 M_2$.
	Due to the triangle inequality, there are two possibilities to place each of points $M_i$, $i=3,4$:

	a) $M_i$ belongs to $l$ and $|M_i M_1|-|M_i M_2| = \pm 1$;

	b) $M_i$ belongs to $m$ and $|M_i M_1| = |M_i M_2|$.

	If (a) is true for both points, then $M\subset l$ and thus $M$ is not an integral point set.
	If (a) is true for one point and (b) is true for another, then the claim of the lemma follows.
	So, suppose to the contrary, that both points $M_3$ and $M_4$ belong to $m$.

>>>>

TODO

@@

\section{Basic Notions and Results}

In this Section, we provide proper definitions and list some known results.

For the sake of brevity, the following designations will be used for numeric sets
of positive integers, non-negative integers and all the integers resp.:
\begin{equation}
	\mathbb{N} = \{1,2,3,4,...\},\quad \mathbb{N}_0 = \mathbb{N} \cup \{0\},
	\quad
	\mathbb{N}_\pm = \{0,\pm 1,\pm 2,\pm 3,\pm 4,...\}
\end{equation}

\begin{lemma}
	\label{lem:triangle_4k_plus_3}
	Any triangle with sides $a \leq b \leq c$, where $c=a+b-1$ and $a, b, c \in \mathbb{N}$,
	has a characteristic of the form $p=4k+3$, $k\in \mathbb{N}_0$.
\end{lemma}

\begin{proof}
	Indeed, let the triangle $ABC$ satisfy: $|BC|=a$, $|AC|=b$. Then, from the triangle inequality $|AB| < |BC|+|AC|$, we can represent the length of side $AB$ as: $|AB|=a+b-s$, $s \in \mathbb{N}$ and $s \leq a$. For $s=1$, the conditions of statement 6 are satisfied: $s$ is an odd number and $s \leq a$.
	Therefore, the triangle with sides $a$, $b$, and $c=a+b-1$ has a characteristic of the form $p=4k+3$, $k\in \mathbb{N}_0$.
\end{proof}
